---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "tools/README-"
)
options(tibble.print_min = 5, tibble.print_max = 5)
```

# rfm 

<!-- badges: start -->
[![CRAN\_Status\_Badge](https://www.r-pkg.org/badges/version/rfm)](https://cran.r-project.org/package=rfm) [![R build status](https://github.com/rsquaredacademy/rfm/workflows/R-CMD-check/badge.svg)](https://github.com/rsquaredacademy/rfm/actions) [![Coverage Status](https://img.shields.io/codecov/c/github/rsquaredacademy/rfm/master.svg)](https://app.codecov.io/github/rsquaredacademy/rfm?branch=master) 
<!-- badges: end -->

## Overview

Tools for customer segmentation using RFM (recency, frequency and monetary) analysis. Generates segments from RFM score. Works with both transaction and customer level data.

## Installation

```{r gh-installation, eval = FALSE}
# Install rfm from CRAN
install.packages("rfm")

# Or the development version from GitHub
# install.packages("pak")
pak::pak("rsquaredacademy/rfm")
```

```{r, echo=FALSE, message=FALSE}
library(rfm)
```

## Usage

**RFM** (recency, frequency, monetary) analysis is a behavior based technique used to segment customers by examining their transaction history such as:

- how recently a customer has purchased (recency)
- how often they purchase (frequency)
- how much the customer spends (monetary)

It is based on the marketing axiom that **80% of your business comes from 20% of your customers**. RFM analysis helps to identify customers who are more likely to respond to promotions by segmenting them into various categories.

To calculate the RFM score we need the following info for each customer: 

- a unique customer id 
- date of transaction/order
- transaction/order amount

```{r rfm_table_order}
# analysis date
analysis_date <- as.Date('2006-12-31')

# generate rfm score
rfm_result <- rfm_table_order(rfm_data_orders, customer_id, order_date,
revenue, analysis_date)

# rfm score
rfm_result

# segment names
segment_names <- c("Champions", "Potential Loyalist", "Loyal Customers",
                   "Promising", "New Customers", "Can't Lose Them",
                   "At Risk", "Need Attention", "About To Sleep", "Lost")

# segment intervals
recency_lower <-   c(5, 3, 2, 3, 4, 1, 1, 1, 2, 1)
recency_upper <-   c(5, 5, 4, 4, 5, 2, 2, 3, 3, 1)
frequency_lower <- c(5, 3, 2, 1, 1, 3, 2, 3, 1, 1)
frequency_upper <- c(5, 5, 4, 3, 3, 4, 5, 5, 3, 5)
monetary_lower <-  c(5, 2, 2, 3, 1, 4, 4, 3, 1, 1)
monetary_upper <-  c(5, 5, 4, 5, 5, 5, 5, 5, 4, 5)

# generate segments
segments <- rfm_segment(rfm_result, segment_names, recency_lower,
recency_upper, frequency_lower, frequency_upper, monetary_lower,
monetary_upper)

segments
```

## Getting Help

If you encounter a bug, please file a minimal reproducible example using 
[reprex](https://reprex.tidyverse.org/index.html) on github. For questions and
clarifications, use [StackOverflow](https://stackoverflow.com/).

