---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# rfm

[![CRAN\_Status\_Badge](http://www.r-pkg.org/badges/version/rfm)](https://cran.r-project.org/package=rfm) [![Travis-CI Build Status](https://travis-ci.org/rsquaredacademy/rfm.svg?branch=master)](https://travis-ci.org/rsquaredacademy/rfm) [![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/rsquaredacademy/rfm?branch=master&svg=true)](https://ci.appveyor.com/project/rsquaredacademy/rfm) [![](https://cranlogs.r-pkg.org/badges/grand-total/rfm)](https://cran.r-project.org/package=rfm)

## Overview

Recency, Frequency and Monetary Value Analysis in R.

## Installation

You can install rfm from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("rsquaredacademy/rfm")
```

```{r, echo=FALSE, message=FALSE}
library(rfm)
```

## Usage

## RFM Table

```{r rfm_table}
analysis_date <- lubridate::as_datetime('2014-04-01 05:30:00', tz = 'UTC')
rfm_result <- rfm_table(rfm_data, customer_id, order_date, revenue, analysis_date)
rfm_result
```

## Heat Map

```{r heatmap}
rfm_heatmap(rfm_result)
```

## Bar Chart

```{r barchart}
rfm_bar_chart(rfm_result)
```

## Scatter Plots

### Monetary Value vs Recency

```{r mr}
rfm_mr_plot(rfm_result)
```

### Frequency vs Monetary Value

```{r fm}
rfm_fm_plot(rfm_result)
```

### Frequency vs Recency

```{r fr}
rfm_fr_plot(rfm_result)
```

Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.
